---
title: "FinalProjectTeam30"
author: "Team 30 - Ron Beiden, Ori Flomin"
date: "2023-05-24"
output: pdf_document
---

**In order to get our  end to end code analysis, run the chunks by the given order **

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE)
 # hide source code in the document
```

##### Loading libraries

```{r load-packages, message = FALSE , include=FALSE,  warning = FALSE}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(tidymodels)
library(randomForest)
library(caret)
library(e1071)
library(nnet)
library(knitr)
library(htmltools)
```


# 1.Loading & Cleaning the Data 

##### Please upload your data here
```{r}
# Load the data and filter to cases from 2011 and after
set.seed(1116)
setwd("C:/Users/1/Desktop/Data_proj/")
final_year <- read.csv("NYPD_Shooting_Incident_Data_2022.csv")
shootings_data <- read.csv("NYPD Shooting Incident - Data 2006-2021.csv") %>%
  mutate(OCCUR_DATE = as.Date(OCCUR_DATE, "%m/%d/%Y")) %>% # Convert OCCUR_DATE column to date format
  filter(OCCUR_DATE >= as.Date("2011-01-01")) # Filter to shootings on or after January 1st, 2011

# Filter data for incidents from 2011 onwards
shootings_data <- shootings_data[shootings_data$OCCUR_DATE >= "2011-01-01", ]
```

##### Handling relevant columns
```{r}
# Filtering the age range into numbers 1-5
shootings_data_set <-function(data){
  data %>% 
  filter(!is.na(PERP_RACE) & PERP_RACE != "UNKNOWN") %>% 
  mutate(PERP_AGE_GROUP_NEW = case_when(
    PERP_AGE_GROUP == "<18" ~ 1,
    PERP_AGE_GROUP == "18-24" ~ 2,
    PERP_AGE_GROUP == "25-44" ~ 3,
    PERP_AGE_GROUP == "45-64" ~ 4,
    PERP_AGE_GROUP == "65+" ~ 5,
    TRUE ~ NA_real_
  ),
  VIC_AGE_GROUP_NEW = case_when(
    VIC_AGE_GROUP == "<18" ~ 1,
    VIC_AGE_GROUP == "18-24" ~ 2,
    VIC_AGE_GROUP == "25-44" ~ 3,
    VIC_AGE_GROUP == "45-64" ~ 4,
    VIC_AGE_GROUP == "65+" ~ 5,
    TRUE ~ NA_real_
  )) %>%
  na.omit()
}
shootings_data_train <- shootings_data_set(shootings_data)
shootings_data_train$OCCUR_TIME_HOUR <- format(as.POSIXct(shootings_data_train$OCCUR_TIME, format="%H:%M:%S"), format="%H")
shootings_data_test <- shootings_data_set(final_year)
shootings_data_test$OCCUR_TIME_HOUR <- format(as.POSIXct(shootings_data_test$OCCUR_TIME, format="%H:%M:%S"), format="%H")
```

```{r}
shootings_data_test <- shootings_data_test %>%
  select(-INCIDENT_KEY, -OCCUR_DATE, -STATISTICAL_MURDER_FLAG) 
shootings_data_train <- shootings_data_train %>%
  select(-INCIDENT_KEY, -OCCUR_DATE, -STATISTICAL_MURDER_FLAG) 
```


# 2.Data Analysis 

##### Summary Statistic

```{r, fig.show="hold", out.width="70%", fig.height=3}
shootings_by_boro <- shootings_data_train %>%
  group_by(BORO) %>%
  summarise(count = n())
ggplot(shootings_by_boro, aes(x=BORO, y=count, fill=BORO)) +
  geom_bar(stat="identity") +
  labs(title="Number of Shootings by Borough", x="Borough", y="Count") +
  scale_fill_manual(values=c("#0072B2", "#E69F00", "#009E73", "#F0E442", "#CC79A7")) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
```{r}
library(ggplot2)

# Group the data by race and count the occurrences for perpetrators, excluding null and empty values
perp_race_counts <- shootings_data %>%
  filter(!is.na(PERP_RACE) & PERP_RACE != "") %>%
  group_by(PERP_RACE) %>%
  summarize(count = n())

# Group the data by race and count the occurrences for victims, excluding null and empty values
vic_race_counts <- shootings_data %>%
  filter(!is.na(VIC_RACE) & VIC_RACE != "") %>%
  group_by(VIC_RACE) %>%
  summarize(count = n())

# Combine the perpetrator and victim race counts into a single data frame, omitting null and empty values
race_counts <- rbind(
  data.frame(Race = perp_race_counts$PERP_RACE, Role = "Perpetrator", Count = perp_race_counts$count),
  data.frame(Race = vic_race_counts$VIC_RACE, Role = "Victim", Count = vic_race_counts$count)
) %>%
  na.omit()

# Create the bar chart with rotated x-axis labels
ggplot(race_counts, aes(x = Race, y = Count, fill = Role)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(x = "Race", y = "Count", fill = "Role") +
  ggtitle("Race Distribution of Perpetrators and Victims in Shooting Incidents") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

ggplot(race_counts, aes(x = Race, y = Count, fill = Role)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(x = "Race", y = "Count", fill = "Role") +
  ggtitle("Race Distribution of Perpetrators and Victims in Shooting Incidents") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

```{r, fig.show="hold", out.width="50%", fig.height=3}
shootings_plot <- function(race_filter = NA, range, step, black_only = FALSE) {
  if (black_only) {
    shootings_data_train_mod <- shootings_data_train[shootings_data_train$PERP_RACE == "BLACK",]
  }
  
  else if (!missing(race_filter)) {
    shootings_data_train_mod <- shootings_data_train %>%
      filter(PERP_RACE != race_filter) %>%
      filter(PERP_RACE != "AMERICAN INDIAN/ALASKAN NATIVE")
  }
  shootings_data_train_mod <- shootings_data_train_mod %>%
    filter(VIC_AGE_GROUP_NEW != 5) %>%
    filter(PERP_AGE_GROUP_NEW != 5)
  ggplot(shootings_data_train_mod, aes(x = PERP_AGE_GROUP_NEW)) + 
    geom_bar(aes(fill = factor(PERP_AGE_GROUP_NEW)), position = "dodge", width = 0.5) + 
    facet_wrap(~PERP_RACE, ncol = 2, scales = "free_x") +
    labs(x = "Perpetrator Age Group", y = "Count", fill = "Victim Age Group") +
    theme_bw() +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
          panel.border = element_blank(), axis.line = element_line(colour = "black"),
          axis.text.x = element_text(angle = -90, hjust = 0)) +
    scale_fill_discrete(name = "Victim Age Group", labels = c("<18", "18-24", "25-44", "45-64")) +
    scale_x_continuous(name = "Perpetrator Age Group", breaks = c(1, 2, 3, 4),
                       labels = c("<18", "18-24", "25-44", "45-64")) +
    scale_y_continuous(breaks = seq(0, 1250, by = step)) +
    theme(legend.position = "bottom") +
    labs(title = "Distribution of Perpetrator and Victim Age Groups by Race",
         subtitle = "Data from shootings in NYC between 2011-2021") +
    coord_cartesian(ylim = c(0, range), expand = FALSE)
  
}
```
```{r, fig.show="hold", out.width="50%", fig.height=4}
shootings_plot(NA, 1250, 250, TRUE)
shootings_plot("BLACK", 300, 100)
```


```{r}
# Heatmap of the geographical locations of incidents based on longitude and latitude
library(leaflet)
library(leaflet.extras)

data1c <- shootings_data_train[c("Latitude", "Longitude")]
data1c <- na.omit(data1c)

incidentsAREA <- leaflet() %>%
  setView(lng = -73.9, lat = 40.7, zoom = 11) %>%
  addProviderTiles("Stamen.Terrain", options = providerTileOptions(minZoom = 2)) %>%
  addHeatmap(data = data1c)
```


### Creating a multi-class classification random forest model to predict the characteristics and patterns of shooting incidents

```{r, message=FALSE}
### Creating a classification model 

# Identify common columns between shootings_data_train and shootings_data_test
common_columns <- intersect(names(shootings_data_train), names(shootings_data_test))

# Refactor columns in shootings_data_test to match the data types of shootings_data_train
for (col in common_columns) {
  col_type_train <- typeof(shootings_data_train[[col]])
  col_type_test <- typeof(shootings_data_test[[col]])

  # Check if the data types are different and refactor if needed
  if (col_type_train != col_type_test) {
    shootings_data_test[[col]] <- as(shootings_data_test[[col]], col_type_train)
  }
}

shootings_data_train <- shootings_data_train  %>%
select(PERP_RACE, PERP_AGE_GROUP, BORO, VIC_AGE_GROUP, VIC_RACE, Latitude, Longitude, BORO)
  
shootings_data_test <- shootings_data_test %>%
  select(PERP_RACE, PERP_AGE_GROUP, BORO, VIC_AGE_GROUP, VIC_RACE, Latitude, Longitude, BORO)

performance <- function(predicator, reference) {
  col_name <- as.character(substitute(predicator))
  reference <- as.character(substitute(reference))
  
  shootings_data_train[[col_name]] <- as.factor(shootings_data_train[[col_name]])
  rf_formula <- as.formula(paste(col_name, "~ PERP_RACE + PERP_AGE_GROUP + BORO + VIC_AGE_GROUP + VIC_RACE + Latitude + Longitude"))
  rf_model1 = randomForest(rf_formula, data = shootings_data_train, ntree = 100)
 
  importances <- importance(rf_model1)
  sorted_importance <- sort(importances, decreasing = TRUE)
  top_10_features <- head(names(sorted_importance), 10)
  predictors <- c(top_10_features, col_name)
  
  shootings_data_train[[col_name]] <- factor(shootings_data_train[[col_name]], ordered = FALSE)
  shootings_data_train[[col_name]] <- relevel(shootings_data_train[[col_name]], ref = reference)
  
  model_output <- capture.output(shootings_model <- multinom(rf_formula, data = shootings_data_train))
  #rm(list = ls())

  test_predict <- predict(shootings_model, newdata = shootings_data_test, "class") %>%
    bind_cols(shootings_data_test %>% select(col_name)) %>%
  rename(.pred_class = 1)
  
  test_predict$.pred_class <- as.character(test_predict$.pred_class)
  
  levels(test_predict$.pred_class) <- levels(shootings_data_test[[col_name]])
  

  tab1 <- table(shootings_data_test[[col_name]], test_predict$.pred_class)

  accuracy <- sum(diag(tab1)) / sum(tab1)
  precision <- diag(tab1) / colSums(tab1)
  recall <- diag(tab1) / rowSums(tab1)
  f1_score <- 2 * precision * recall / (precision + recall)
  
  # Get the maximum values (assuming binary classification)
  accuracy <- max(accuracy, na.rm = TRUE)
  precision <- max(precision, na.rm = TRUE)
  recall <- max(recall, na.rm = TRUE)
  f1_score <- max(f1_score, na.rm = TRUE)

  # Print the results
  cat(sprintf("Accuracy: %.2f\n", accuracy))
  cat(sprintf("Precision: %.2f\n", precision))
  cat(sprintf("Recall: %.2f\n", recall))
  cat(sprintf("F1 Score: %.2f\n", f1_score))
  
  return(rf_model1)
}
```

### Calculating and displaying the likelihood probabilities of the shooting incidents

```{r, message=FALSE}
calculate_mean <- function(rf_model2, varr) {
  
  predict_probs <- predict(rf_model2, type = "prob")
  column <- match(varr, colnames(predict_probs))

    if (is.na(column)) {
    stop("var not found in the matrix.")
  }
  
  values <- predict_probs[, column]
  numeric_values <- as.numeric(values)
  mean_value <- mean(numeric_values, na.rm = TRUE)
  
  return(mean_value)
}
# Displaying data on the probability of different groups being the race of the perpetrator or the borough where the incidents occur

model_input <- performance(BORO, "BRONX")
calculate_mean(model_input, "BRONX")
calculate_mean(model_input, "BROOKLYN")
calculate_mean(model_input, "MANHATTAN")

model_input <- performance(PERP_RACE, "BLACK")
calculate_mean(model_input, "BLACK")
calculate_mean(model_input, "WHITE")
calculate_mean(model_input, "WHITE HISPANIC")
```